---
title: "Tutorial"
author: "Tobias K. Mildenberger"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
  pdf_document:
    number_sections: false
    toc: false
    keep_tex: false
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "  ", fig.align = 'center', cache=FALSE)
```

# Description
This package allows to process scientific survey data and fit spatio-temporal
fish distribution models to the processed data. This tutorial highlights the
main functionality of the package from loading the package to fitting the
distribution models.


# Load the package
The R package is hosted on GitHub and can be installed using the function
`install_github` from the `remotes` R package with:

```{r, eval = FALSE}
remotes::install_github("tokami/fishdish/fishdish")
```

After the successful installation, the package can be loaded into R with:

```{r, eval = FALSE}
library(fishdish)
```


# Required data

## Download the DATRAS data set
In order to download the DATRAS data set which contains the information of the
scientific surveys, we can use the function `download.data`. Note, that the
following code downloads the whole DATRAS database which can take several
minutes. The arguments of the function allow you to limit the download to the
required years, surveys, quarters, or species (defined by `aphiaID`). However,
note that already selecting one (or more) species at this step, might affect the
number of hauls with zero observations later on.

```{r, eval = FALSE}
datras <- download.data(first.year = 1967, last.year = 2021, surveys = "all",
                        quarters = "all", aphiaID = "all", datasets = c("HH","HL","CA"),
                        calc.swept.area = FALSE,
                        reduce.file.size = FALSE)
```

```{r, eval = FALSE, echo = FALSE}
load("../../data/datras.RData")
```


## Download the bathymetric data set
As depth is an important predictor of fish distributions, we need to download a
bathymetric map. This could be done with the function `getNOAA.bathy` in the
`marmap` package. Make sure to download depth information for the whole area of
interest and adjust the grid size as required. This depth grid will be used
later for the definition of the prediction grid.

The data directory at [https://github.com/tokami/fishdish/data] includes an
example bathymetric map for the Northeast Atlantic with a grid size adjusted to
the ICES statistical subrectangles (20' x 10' resolution). After downloading it,
it can be loaded into R with:

```{r, eval = FALSE}
load("bathy.sub.statrec.grid.RData")
```

```{r, eval = FALSE, echo = FALSE}
load("../../data/bathy.sub.statrec.grid.RData")
```


# Define species for analysis
The DATRAS data set include information about more than 1000 species, if you do
not want to analyse all of them, define a vector with the Aphia IDs of the
species. The species names are not needed, but might be helpful.

```{r}
## 4 example species
species <- c(
    "Gadus morhua" = 126436,
    "Merluccius merluccius" = 126484,
    "Scophthalmus maximus" = 127149,
    "Conger conger" = 126285
)
```



# Define a prediction grid for each species
In theory, we could use the same prediction grid for all species, however, it
might be meaningful to restrict the extent of extrapolation by having
species-specific prediction grids. These grids could be constant over time or
time-variant accounting for the fact that the various surveys started operating
at different times and, thus, the area sampled changed over time. Here, we
define time-variant prediction grids for each species by identifying the ICES
areas where a given species was caught at least once in a given year. If a
species was observed at least once in a given ICES area, the whole area is
included into the prediction grid for this year.


```{r, eval = FALSE}
## Create an empty list
pred.grid.fine <- vector("list", length(species))

## Loop through all species
for(i in 1:length(species)){

    ## Extract data set by species
    dati <- prep.species(prep.data(datras, AphiaID = species[i], est.n = TRUE, est.bio = TRUE),
                         min.gears = 0, min.surveys = 0, min.hauls = 0)

    ## Define prediction grid for each year
    years <- unique(dati$Year)

    ## Make an empty list
    pred.grid.fine[[i]] <- vector("list", length(years))

    ## Loop through all years
    for(y in 1:length(years)){

        ## Subset data set for specific year
        datiy <- dati[which(as.numeric(as.character(dati$Year)) == years[y]),]

        ## Make sure to drop other years if year is a factor
        if(inherits(datiy$Year, "factor")) datiy$Year <- droplevels(datiy$Year)

        ## Use some criteria to extract information from bathymetric data set
        ind <- which(bathy.sub.statrec.grid$Area_27 %in% unique(datiy$Area_27))
        pred.grid.fine[[i]][[y]] <- bathy.sub.statrec.grid[ind,]

        ## Rename the area to SweptArea (eases predicting absolute abundances)
        colnames(pred.grid.fine[[i]][[y]])[colnames(pred.grid.fine[[i]][[y]]) == "SubStatRec_area"] <- "SweptArea"

        ## Add a column with year (eases merging if required later)
        pred.grid.fine[[i]][[y]]$Year <- as.numeric(as.character(years[y]))
    }

    ## Add names of years as names of prediction grids
    names(pred.grid.fine[[i]]) <- as.character(years)

}

## Add species names as names of list with prediction grids
names(pred.grid.fine) <- gsub(" ", "_", names(species))
```

If instead of using the whole ICES area, you would like to use a statistical
rectangle you could replace the code `ind <-
which(bathy.sub.statrec.grid$Area_27 %in% unique(datiy$Area_27))` with `ind <-
which(bathy.sub.statrec.grid$StatRec %in% unique(datiy$StatRec))`.



# Prepare the data sets
The DATRAS data set requires some processing in order to fit spatial or
spatio-temporal distribution models to the data. For example, it is necessary to
identify the real zero observations, i.e. the hauls that could have caught the
species, but did not. Furthermore, the following functions allow to estimate the
number and biomass of all individuals caught in a specific haul or for juveniles
and adults separately.


```{r, eval = FALSE}
## Create empty list
data.list <- vector("list", length(species))

## Loop through all species
for(i in 1:length(species)){

    ## Prepare data set
    dat.spec <- prep.data(datras, AphiaID = species[i], est.n = TRUE, est.bio = TRUE, split.juv.adults = TRUE)

    ## Prepare species datas
    data.list[[i]] <- prep.species(dat.spec, min.gears = 0, min.surveys = 0, min.hauls = 0)
}

## Add species names as names of data list
names(data.list) <- gsub(" ", "_", names(species))
```




# Predict the species distribution using a spatio-temporal GAM
Any spatio-temporal distribution model could now be fitted to the processed
species data set (elements of the `data.list` list) and be used to predict the
abundance or relative density for the prediction grid (elements of the
`pred.grid.fine` list). The models included in this package use the
spatio-temporal generalised additive models (GAMs) of the `surveyIndex` package
by Casper W. Berg.

In order to fit these GAMs, just select the data set for the species of interest
and define which life stage to use by overwriting N. If required, you could
subset the data set to include any years of interest with the following code.
For example, for juvenile cod and observations after 2009, we can use:


```{r, eval = FALSE}
## Subset the data set for specific species
dati.spec <- data.list[[1]]

## Define the numbers to be used in column 'N',
## e.g. to model the distribution of the juveniles:
dati.spec$N <- dati.spec$n.juv

## Subset years
dati.spec <- subset(dati.spec, Year > 2009)
```

Next, we have to define the model structure of the GAM. A simple structure could
for example be:

```{r, eval = FALSE}
mod <-  "s(lon, lat, bs=c('ds'), k=128, m=c(1,0.5)) + s(ctime, bs='ds', k=17, m=c(1,0)) + ti(ctime, lon, lat, d=c(1,2), bs=c('ds','ds'), k=c(17,5), m=list(c(1,0), c(1,0.5))) + te(timeOfYear, lon, lat, d=c(1,2), bs=c('cc','ds'), k=c(6,30), m=list(c(1,0), c(1,0.5))) + s(Depth, bs='ds', k=5, m=c(1,0)) + offset(log(SweptArea))"
```

Note, that this model does not include any gear or ship-gear effect which is
likely to be important when combining information from many different surveys.
You can get a suggestion for meaningful model structures with the function
`list.recom.models(dati.spec)`.

Finally, we can predict the fish distribution with the function `est.dist.one`.

```{r, eval = FALSE}
## Fit GAM
fit <- est.dist.one(dati.spec, mod,
                    grid = pred.grid.fine[[1]],
                    verbose = FALSE)
```

The argument `predfix` allows to specify specific prediction parameters and
could be used in the following way:


```{r, eval = FALSE}
## Prediction parameters
tmp <- subset(dati.spec, N > 0)
predfix <- list()

## First quarter
predfix$timeOfYear <- median(tmp$timeOfYear[tmp$Quarter == 1])

## Most common gear
predfix$Gear <- names(which.max(table(tmp$Gear)))[1]

## Median swept area index
predfix$SweptArea <- median(tmp$SweptArea)

## Most common ship-gear interaction
predfix$ShipG <- names(which.max(table(tmp$ShipG)))[1]

## Fit GAM with predfix
fit <- est.dist.one(dati.spec, mod,
                    grid = pred.grid.fine[[1]],
                    predfix = predfix,
                    verbose = FALSE)
```

The results can then be plotted with the various plotting functions included in
the package. For example, to plot the abundance over time or the distribution in
space, the following functions can be used.

```{r, eval = FALSE}
## Plot the estimated abundance
plotfishdish.abun(fit)

## Plot the estimated distribution
plotfishdish.dist(fit)
```
